Identifying and correcting for misspecifications\\in GWAS summary statistics and polygenic scores

### 2022-06-30 paper/paper-misspec.tex
red markup

Here we investigate some of these misspecifications, and propose adjustments to improve the predictive performance of polygenic scores derived from GWAS summary statistics.
We approach this from three different angles.
First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite[]{prive2020ldpred2}.
This QC consists in comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel.
This is useful to check that the input parameters used are consistent with one another. 
This was particularly important for LDpred2-auto, which directly estimates two key model parameters, the SNP heritability and polygenicity, from the data \cite[]{prive2020ldpred2}.
Here we further show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
Second, we investigate possible adjustments to apply to the input parameters of PGS methods using summary statistics, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and corresponding sample sizes.
For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude than when computed from true genotypes.
Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
Note that, in this paper, we use ``robust'' to mean ``similar predictive performance''. 
We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.
We focus our investigations on LDpred2 and lassosum2 for two reasons. First, multiple studies have shown that LDpred2 and lassosum rank among the best methods for single-trait polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
Second, lassosum2 now uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
We additionally investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats required for the LD matrices and input GWAS summary statistics.

@@

We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020). These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.  
This results in 362,307 individuals of Northwestern European ancestry.
We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).

@@

We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals of Northwestern European ancestry as described in the previous section. We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.

@@

\subsection{Model for summary statistics}

@@

It is assumed that the marginal (GWAS) effects follow
\begin{align}
\begin{split}\label{eq:model}
 \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(S^{-1} R S \gamma, S^{-1} R S^{-1}) ~,\\
 S \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(R S \gamma, R) ~.
\end{split}
\end{align}
where $R$ is the correlation matrix of genotypes \cite[]{zhu2017bayesian}. 
Then PGS methods aim at inferring $\gamma$, the true causal (per-allele) effects, from $S$, $R$, and $\hat{\gamma}$.
However, in practice, we only have estimates for $S$ and $R$, which causes some model misspecifications.

@@

We can impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
\begin{equation}\label{eq:imputeN}
n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
\end{equation}
and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
\begin{equation}\label{eq:imputeNeff}
n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
\end{equation}
In practice, we also bound these estimates to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile (robust minimum) of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$.

@@

Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net). 
Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
\[
\beta_j^{(t)} =
\begin{cases}
\dfrac{\text{sign}\big(u_j^{(t)}\big) \big( \big|u_j^{(t)}\big| - \lambda \big)}{\widetilde{X}_j^T \widetilde{X}_j + \delta} & \text{if } \big|u_j^{(t)}\big| > \lambda ~, \\
0 & \text{otherwise.}
\end{cases}
\]
where 
\[
u_j^{(t)} = r_j - \widetilde{X}_j^T \big( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \big) ~.
\]
Following the notations from \cite{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$. Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, then update this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only. Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$. 
Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.

In this new implementation of the lassosum model, which we call lassosum2, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$. 
Therefore, except for the two hyper-parameters, lassosum2 uses the exact same input parameters as LDpred2 \cite[]{prive2020ldpred2}.
We try $\delta \in \{0.001, 0.01, 0.1, 1\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001. By default in lassosum2, we use a similar sequence of 30 values, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.
Note that we do not provide an ``auto'' version using pseudo-validation (as in lassosum \cite[]{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure \ref{fig:pseudoval}).
Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
Contrary to for lassosum, we do not require splitting the genome into independent LD blocks anymore, yet we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).

@@

We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.

@@

We design GWAS simulations where variants have different sample sizes, which is often the case when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
We then randomly divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter. 
We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto, SBayesR, and PRS-CS-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.160 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, 0.141 for LDpred2-inf, 0.138 for SBayesR, and 0.159 for PRS-CS-auto (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
Results when using imputed sample sizes are very similar to when using the true ones.
Note that C+T does not use this sample size information, and that PRS-CS can only be provided with a single value (we use the maximum one).
When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.158 for lassosum and to 0.164 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.122 for LDpred2-auto, 0.125 for LDpred2-inf (Figure \ref{fig:simu-misN}).
This extreme simulation scenario shows that LDpred2 can be sensitive to the misspecification of per-variant GWAS sample sizes, whereas lassosum (and lassosum2) seems little affected by this.
With 0.141 for SBayesR, predictive performance is actually larger than when using the true per-variant sample sizes.

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario. Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean. The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T). Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}), which is a requirement for lassosum and PRS-CS.}
	\label{fig:simu-misN}
\end{figure}

We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal $p$ estimate of LDpred2-auto (and similarly the proportion of null variants in SBayesR) always continuously increases up to $1$, which makes them behave as an infinitesimal model ($p = 1$).
To overcome this limitation in LDpred2-auto, we introduce two new parameters to make it more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
In these simulations with misspecified per-variant sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}). 
Indeed, using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
Here we call ``LDpred2-low-h2'' when running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$.
Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.164 when using the maximum sample size.
Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods section \ref{sec:LD}). This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).

As secondary analysis, we rerun this simulation with a smaller heritability of 4\% instead. While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure \ref{fig:simu-misN-smallh2}).
We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained from LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).

@@

We use other external GWAS summary statistics for which INFO scores are reported (see Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for T1D-affy (0.885).
QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
We recall that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.
In figures \ref{fig:res_cad} and \ref{fig:res_brca_onco}-\ref{fig:res_t1d_aff}, we then provide similar results as figure \ref{fig:res_brca_icogs} for other phenotypes.
For MDD and PRCA, most changes introduced before have little to no impact on predictive performance.
For CAD and T1D, the QC proposed in \cite{prive2020ldpred2} and the additional QC proposed here provides much better predictive performance, especially for LDpred2-auto (and LDpred2-auto-rob) than when using no quality control, showing how important this preliminary step is.

>>>>

Here we investigate some of these misspecifications, and propose adjustments to improve the predictive performance of polygenic scores derived from GWAS summary statistics.
We approach this from three different angles.
First, based on additional summary information such as the imputation INFO scores and allele frequencies from the GWAS summary statistics, we refine our previously proposed QC \cite[]{prive2020ldpred2}.
This QC consists in comparing standard deviations (of genotypes) inferred from GWAS summary statistics with the ones computed from a reference panel.
This is useful to check that the input parameters used are consistent with one another. 
This was particularly important for LDpred2-auto, which directly estimates two key model parameters, the SNP heritability and polygenicity, from the data \cite[]{prive2020ldpred2}.
Here we further show that standard deviations of imputed genotypes (allele dosages) are lower than the expected values under Hardy--Weinberg equilibrium.
Second, we investigate possible adjustments to apply to the input parameters of PGS methods using summary statistics, namely the reference LD (linkage disequilibrium) matrix, the GWAS effect sizes, their standard errors and corresponding sample sizes.
For example, we show that GWAS effect sizes computed from imputed dosages are larger in magnitude than when computed from true genotypes.
Third, we introduce two new optional parameters in LDpred2-auto to make it more robust to these types of misspecification.
{\color{red}Note that, in this paper, we use ``robust'' to mean ``similar predictive performance''. 
We also reimplement and use a new version of lassosum, called lassosum2, in which we better handle when per-variant sample sizes are different.}
We focus our investigations on LDpred2 and lassosum2 for two reasons. First, multiple studies have shown that LDpred2 and lassosum rank among the best methods for single-trait polygenic prediction \cite[]{mak2017polygenic,prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
Second, lassosum2 now uses the exact same input parameters as LDpred2, which makes it easy for us to test the different QCs and adjustments presented here.
{\color{red}We additionally investigate PRS-CS and SBayesR, two competitive PGS methods \cite[]{ge2019polygenic,lloyd2019improved}, for which we develop some code to convert between the different formats required for the LD matrices and input GWAS summary statistics.}

@@

We use the UK Biobank imputed (BGEN) data \cite[]{bycroft2018uk}. 
We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020). These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \cite{bycroft2018uk}).
To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs and further restrict individuals to those within a log-distance of 5 \cite[]{prive2020efficient}.  
This results in 362,307 individuals {\color{red}of Northwestern European ancestry}.
We randomly sample 300,000 individuals to form a training set (e.g.\ to run the GWAS), 10,000 individuals to form a validation set (to tune hyper-parameters), and use the remaining 52,307 individuals to form a test set (to evaluate final predictive models).

@@

We also use the UK Biobank data for validation/testing in real data analyses, and use the same individuals {\color{red}of Northwestern European ancestry} as described in the previous section. We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
We restrict to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.

@@

{\color{red}\subsection{Model for summary statistics}}

@@

{\color{red}It is assumed that the marginal (GWAS) effects follow
\begin{align}
\begin{split}\label{eq:model}
 \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(S^{-1} R S \gamma, S^{-1} R S^{-1}) ~,\\
 S \hat{\gamma} | S,R,\gamma & \sim \mathcal{N}(R S \gamma, R) ~.
\end{split}
\end{align}
where $R$ is the correlation matrix of genotypes \cite[]{zhu2017bayesian}. 
Then PGS methods aim at inferring $\gamma$, the true causal (per-allele) effects, from $S$, $R$, and $\hat{\gamma}$.
However, in practice, we only have estimates for $S$ and $R$, which causes some model misspecifications.
Moreover, ubiquitous correlations between the large number of variants may cause a Gibbs sampler or any iterative approach used in PGS methods to fail properly estimating genetic effect sizes \cite[]{roberts1997updating}.}

@@

We can impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
\begin{equation}\label{eq:imputeN}
n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
\end{equation}
and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
\begin{equation}\label{eq:imputeNeff}
n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
\end{equation}
In practice, we also bound these estimates to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the total (effective) sample size.
Also note that the estimate of $\text{var}(G_j)$ has to account for the imputation accuracy, i.e.\ use $2 \cdot f_j \cdot (1 - f_j) \cdot \text{INFO}_j$ (Section \ref{sec:imputation}).
{\color{red}In equations \eqref{eq:approx-sd-lin} and \eqref{eq:imputeN}, we estimate $\text{var}(y)$ by the first percentile (robust minimum) of $0.5 \left(N ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2\right)$.}

@@

Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$ (where $0 < s \leq 1$), we use $R_{\delta} = R + \delta I$ (where $\delta > 0$), which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net). 
Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
\[
\beta_j^{(t)} =
\begin{cases}
\dfrac{\text{sign}\big(u_j^{(t)}\big) \big( \big|u_j^{(t)}\big| - \lambda \big)}{\widetilde{X}_j^T \widetilde{X}_j + \delta} & \text{if } \big|u_j^{(t)}\big| > \lambda ~, \\
0 & \text{otherwise.}
\end{cases}
\]
where 
\[
u_j^{(t)} = r_j - \widetilde{X}_j^T \big( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \big) ~.
\]
Following the notations from \cite{prive2020ldpred2}, denote $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$. Moreover, using the notations from \cite{prive2020ldpred2}, $u_j^{(t)} = \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} + \beta_j^{(t-1)}$,
where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
Then computing $R_{.,j}^T \beta^{(t-1)}$ is the most time-consuming part of each iteration.
To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with an initial vector of 0s only (for all $j$) since $\beta^{(0)} \equiv 0$, then update this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only. Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$. 
{\color{red}Since bigsnpr v1.10.4, we now also use this updating strategy in LDpred2 (-grid and -auto), which makes it much faster, especially when the polygenicity is small.}

In this new implementation of the lassosum model, which we call lassosum2, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$. 
Therefore, except for the two hyper-parameters, lassosum2 uses the exact same input parameters as LDpred2 \cite[]{prive2020ldpred2}.
We try $\delta \in \{0.001, 0.01, 0.1, 1\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001. By default in lassosum2, we use a similar sequence of 30 values, but between $\lambda_0$ and $\lambda_0 / 100$ instead, where $\lambda_0 = \max_j \big|\widehat{\beta}_j\big|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
{\color{red}To make lassosum2 more robust to different per-variant sample sizes $n_j$, we also introduce per-variant penalty factors $\sqrt{\max_k(n_k) / n_j}$, by which we internally multiply $\lambda$ and $\delta$ to penalize variants with smaller GWAS sample sizes more.}
Note that we do not provide an ``auto'' version using pseudo-validation (as in lassosum \cite[]{mak2017polygenic}) since we have not found the pseudo-validation scores to be consistent enough with the predictive performance (Figure \ref{fig:pseudoval}).
Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated.
Contrary to for lassosum, we do not require splitting the genome into independent LD blocks anymore, yet we recommend to do so for robustness in LDpred2, and for extra speed gain (cf.\ Methods section \ref{sec:LD}).

@@

We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
{\color{red}Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).}
Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
{\color{red}Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.}

@@

We design GWAS simulations where variants have different sample sizes, which is often the case when meta-analyzing GWAS summary statistics from multiple cohorts with different sets of variants \cite[]{wang2021global}.
Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes with a heritability of 20\% and 2000 causal variants.
We then randomly divide the 40,000 variants into three groups: for half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of the variants and 60\% for the remaining quarter. 
We then run C+T, lassosum, lassosum2 (Methods section \ref{lassosum2}), LDpred2-inf, LDpred2(-grid), LDpred2-auto, SBayesR, and PRS-CS-auto \cite[]{prive2019making,mak2017polygenic,prive2020ldpred2,lloyd2019improved,ge2019polygenic} by using either the true per-variant GWAS sample sizes, the total sample size for all variants, or per-variant sample sizes imputed using equation \eqref{eq:imputeN}.
When providing true per-variant GWAS sample sizes, squared correlations between the polygenic scores and the simulated phenotypes are of 0.123 for C+T, 0.160 for lassosum, 0.169 for lassosum2, 0.159 for LDpred2(-grid), 0.140 for LDpred2-auto, 0.141 for LDpred2-inf, 0.138 for SBayesR, and 0.159 for PRS-CS-auto (Figure \ref{fig:simu-misN}, averaging over 10 simulations).
Results when using imputed sample sizes are very similar to when using the true ones.
Note that C+T does not use this sample size information{\color{red}, and that PRS-CS can only be provided with a single value (we use the maximum one)}.
When using the total GWAS sample size instead of the per-variant sample sizes, predictive performance slightly decreases to 0.158 for lassosum and to 0.164 for lassosum2, but dramatically decreases for LDpred2 with new values of 0.134 for LDpred2-grid, 0.122 for LDpred2-auto, 0.125 for LDpred2-inf (Figure \ref{fig:simu-misN}).
This extreme simulation scenario shows that LDpred2 can be sensitive to the misspecification of per-variant GWAS sample sizes, whereas lassosum (and lassosum2) seems little affected by this.
{\color{red}With 0.141 for SBayesR, predictive performance is actually larger than when using the true per-variant sample sizes.}

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
	\caption{Results for the simulations with sample size misspecification, averaged over 10 simulations for each scenario. Reported 95\% confidence intervals are computed from 10,000 non-parametric bootstrap replicates of the mean. The GWAS sample size is either ``true'' when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants, ``imputed'' (cf.\ equation \eqref{eq:imputeN}), or ``any'' when the method does not use this information (the case for C+T). Red bars correspond to using the LD with independent blocks (Methods section \ref{sec:LD}){\color{red}, which is a requirement for lassosum and PRS-CS}.}
	\label{fig:simu-misN}
\end{figure}

We conduct further investigations to explain the results of figure \ref{fig:simu-misN}.
{\color{red}First, the results for LDpred2-auto and SBayesR are similar to LDpred2-inf in these simulations because the internal $p$ estimate of LDpred2-auto (and similarly the proportion of null variants in SBayesR) always continuously increases up to $1$, which makes them behave as an infinitesimal model ($p = 1$).}
To overcome this limitation in LDpred2-auto, we introduce two new parameters to make it more robust, and refer to this as ``LDpred2-auto-rob'' here (Methods section \ref{sec:robust}).
The first parameter prevents $p$ from diverging to $1$, while the second shrinks the off-diagonal elements of the LD matrix (a form of regularization). 
Second, for lassosum2, results for a grid of parameters (over $\lambda$ and $\delta$) are quite smooth compared to LDpred2 (Figures \ref{fig:lassosum2-misN} and \ref{fig:ldpred2-misN}). 
In these simulations with misspecified per-variant sample sizes, it seems highly beneficial to use a small value for the SNP heritability hyper-parameter $h^2$ in LDpred2, e.g.\ a value of 0.02 or even 0.002 when the true value is 0.2 (Figure \ref{fig:ldpred2-misN}). 
Indeed, using a small value for this hyper-parameter induces a larger regularization (shrinkage) on the effect sizes.
Here we call ``LDpred2-low-h2'' when running LDpred2(-grid) with a grid of hyper-parameters including these low values for $h^2$.
Results with LDpred2-low-h2 improves from 0.159 to 0.169 when using true sample sizes and from 0.134 to 0.164 when using the maximum sample size.
Finally, we introduce a last change for robustness here: we form independent LD blocks in the LD matrix to prevent small errors in the Gibbs sampler to propagate to too many variants (Methods section \ref{sec:LD}). This change seems to solve convergence issues of LDpred2 in these simulations (Figure \ref{fig:ldpred2-misN}) and further improves predictive performance for all LDpred2 methods (Figure \ref{fig:simu-misN}).

{\color{red}As secondary analysis, we rerun these simulations with a smaller heritability of 4\% instead. While predictive performance are overall much lower, relative performance are less attenuated when using the maximum sample size, otherwise results are highly consistent as before (Figure \ref{fig:simu-misN-smallh2}).
We also try using the shrunk LD matrix from GCTB that is used e.g.\ for SBayesR \cite[]{lloyd2019improved}; in these simulations, higher predictive performance are obtained from LDpred2-auto with this shrunk LD matrix, whereas SBayesR benefit from using LDpred2's windowed LD matrix (Figure \ref{fig:simu-misN-shrunk}).}

@@

We use other external GWAS summary statistics for which INFO scores are reported (see Table \ref{tab:sumstats}); they all have a very high mean INFO score (larger than 0.94), except for T1D-affy (0.885).
QC plots comparing standard deviations usually show little deviation from the identity line (after the INFO score correction), except for coronary artery disease (CAD) summary statistics (Figures \ref{fig:brca_icogs_qc}-\ref{fig:t1d_illu_qc}).
{\color{red}Note that the popular CAD summary statistics used here come from a multi-ancestry GWAS meta-analysis with more than 20\% non-European samples \cite[]{nikpay2015comprehensive,prive2021using}.}
In figures \ref{fig:res_cad} and \ref{fig:res_brca_onco}-\ref{fig:res_t1d_aff}, we then provide similar results as figure \ref{fig:res_brca_icogs} for other phenotypes.
For MDD and PRCA, most changes introduced before have little to no impact on predictive performance.
For CAD and T1D, the QC proposed in \cite{prive2020ldpred2} and the additional QC proposed here provides much better predictive performance, especially for LDpred2-auto (and LDpred2-auto-rob) than when using no quality control, showing how important this preliminary step is.
